---
title: "Housing Affordability Across Ontario Municipalities"
author: "Kieran Shah"
date: "September 14, 2026"

format:
  beamer:
    aspectratio: 169
    slide-level: 2
    keep-tex: false

execute:
  echo: false
  warning: false
  message: false

params:
  data_file: "data/processed/final_analytic_dataset.csv"
  output_dir: "eda/"
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)

knitr::opts_knit$set(
  root.dir = "H:/github_repos/Assignment_MMAH/"
)

library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(scales)
library(sf)
library(ggplot2)
library(scales)
library(knitr)
library(stringr)
library(leaflet)

analytic_file <- "H:/github_repos/Assignment_MMAH/data/processed/final_analytic_dataset.csv"

# Point this to the .shp file you downloaded
boundary_file <- "C:/Users/shahki/Downloads/lcsd000b21a_e/lcsd000b21a_e.shp"


# ------------------------------------------------------------
# 2. Read analytic data
# ------------------------------------------------------------

analytic_data <- fread(
  analytic_file,
  encoding = "Latin-1"
) %>%
  as_tibble() %>%
  dplyr::rename(
    population_2021 =
      `Population, 2021`,
    population_growth_pct =
      `Population percentage change, 2016 to 2021`,
    population_density =
      `Population density per square kilometre`,
    median_age =
      `Median age of the population`,
    average_household_size =
      `Average household size`,
    median_income_1p =
      `Median total income of one-person households in 2020 ($)`,
    renter_pct =
      `Renter`,
    unsuitable_major_repairs_pct =
      `'Not suitable' and 'major repairs needed'`,
    housing_target_status =
      `Housing Target Status`
  ) %>% 
  mutate(
    
    # Extract 7-digit Census Subdivision identifier
    CSDUID = str_sub(
      as.character(DGUID),
      -7
    ),
    
    # Calculate affordability measures
    affordability_30plus =
      100 *
      (`30% to less than 50%` + `50% or more`) /
      `Total - Shelter-cost-to-income ratio`,
    
    affordability_50plus =
      100 *
      `50% or more` /
      `Total - Shelter-cost-to-income ratio`,
    affordability_30plus_pct =
      100 * (`30% to less than 50%` + `50% or more`) / `Total - Shelter-cost-to-income ratio`,
    
    affordability_50plus_pct =
      100 * `50% or more` / `Total - Shelter-cost-to-income ratio`,
    
    log_population_density =
      log1p(population_density),
    
    # A municipality name in the CMHC join indicates actual CMHC coverage.
    # Rows without a CMHC match may currently contain zero after NA replacement;
    # restore these to NA so "not covered" is not interpreted as "zero starts".
    cmhc_covered =
      !is.na(municipality),
    
    starts_2020_cmhc =
      if_else(
        cmhc_covered,
        as.numeric(starts_2020),
        NA_real_
      ),
    
    starts_2021_cmhc =
      if_else(
        cmhc_covered,
        as.numeric(starts_2021),
        NA_real_
      ),
    
    starts_2020_per_1000_clean =
      if_else(
        cmhc_covered &
          !is.na(population_2021) &
          population_2021 > 0,
        1000 * starts_2020_cmhc / population_2021,
        NA_real_
      )
  )


```

## Introduction and Purpose

:::: small-text
::: small-text
Affordability is major concern across Ontario.

- A recent poll conducted by Angus Reid identified the cost of living and housing affordability as a the most important issue.

This presentation will seek to **identify which factors are most associated with affordability** using Statistics Canada data and a regression model.

Statistics Canada provided the two main data sources for this presentation at the **Census Subdivision level (CSD)**:

1.  `Table 98-10-0247-01` provided affordability variables.
2.  `Table 98-401-X2021021` is the summary of the predictor variables.
:::
::::

## EDA: Affordability Outcome Distribution {style="{.smaller}"}

```{r eda_1}


mean_values <- tibble(
  Measure = c("Spending ≥30%", "Spending ≥50%"),
  mean_value = c(14.8, 4.8)
)

analytic_data %>%
  select(
    `Spending ≥30%` = affordability_30plus_pct,
    `Spending ≥50%` = affordability_50plus_pct
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Measure",
    values_to = "Percent"
  ) %>%
  ggplot(aes(x = Percent)) +
  geom_histogram(
    bins = 25,
    boundary = 0
  ) +
  geom_vline(
    data = mean_values,
    aes(xintercept = mean_value),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_text(
    data = mean_values,
    aes(
      x = mean_value,
      y = Inf,
      label = paste0("Mean = ", mean_value, "%")
    ),
    vjust = 1.5,
    hjust = -0.1,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~ Measure,
    nrow = 1,
    scales = "free_y"
  ) +
  labs(
    x = "Households (%)",
    y = "CSDs",
    title = "Distribution of housing affordability (n=577, missing=168)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

```

## Affordability Map

```{r mapping_slide}

csd <- st_read(
  boundary_file,
  quiet = TRUE
)

ontario_csd <- csd %>%
  filter(PRUID == "35") %>%
  mutate(
    CSDUID = as.character(CSDUID)
  )

map_data <- ontario_csd %>%
  left_join(
    analytic_data,
    by = "CSDUID"
  )

map_data_static <- map_data %>%
  st_make_valid() %>%
  st_transform(3347)

ggplot(map_data_static) +
  geom_sf(
    aes(fill = affordability_30plus),
    color = "white",
    linewidth = 0.08
  ) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey85",
    name = "Households\nspending ≥30%"
  ) +
  coord_sf(datum = NA) +
  labs(
    title = "Housing affordability burden across Ontario"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

## EDA Predictors

```{r eda_pred}
#| fig-height: 3.5
#| fig-width: 7.5
#| out-width: "95%"
univariate_plot_data <- analytic_data  %>% 
  transmute(
    affordability_30plus,
    `Renter share (%)` = renter_pct,
    `Median age (years)` = median_age,
    `Median income ($)` = median_income_1p,
    `Population Density` = population_density,
    `Log population density` = log_population_density,
    `Average household size` = average_household_size,
    `Unsuitable + major repairs (%)` = unsuitable_major_repairs_pct,
    `Population growth 2016–2021 (%)` = population_growth_pct
  ) %>% 
  pivot_longer(
    cols = -affordability_30plus,
    names_to = "Predictor",
    values_to = "Value"
  ) %>% 
  filter(
    !is.na(affordability_30plus),
    !is.na(Value)
  )

univariate_correlations <- univariate_plot_data  %>% 
  group_by(Predictor)  %>% 
  summarise(
    r = cor(
      Value,
      affordability_30plus,
      use = "complete.obs"
    ),
    .groups = "drop"
  ) %>% 
  mutate(
    facet_label = paste0(
      Predictor,
      "\nr = ",
      sprintf("%.2f", r)
    )
  )

univariate_plot_data <- univariate_plot_data  %>% 
  left_join(
    univariate_correlations,
    by = "Predictor"
  )

ggplot(
  univariate_plot_data,
  aes(
    x = Value,
    y = affordability_30plus
  )
) +
  geom_point(
    alpha = 0.45,
    size = 1.2
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 0.7
  ) +
  facet_wrap(
    ~ facet_label,
    scales = "free_x",
    ncol = 3
  ) +
  labs(
    x = NULL,
    y = "Households spending ≥30% (%)",
    title = "Univariate relationships with housing affordability burden"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(
      face = "bold",
      size = 9
    ),
    plot.title = element_text(
      face = "bold"
    )
  )

```

## Modelling Approach {style="{.smaller}"}

```{r}
predictor_summary <- analytic_data %>% 
  mutate(
    
    density_100 =
      population_density / 100,
    
    log_density =
      log1p(population_density),
    income_10k =
      median_income_1p / 10000
    
  ) %>% 
  rename(
    `Percent Renting` =renter_pct,
    `Median Age` = median_age,
    `Median Total Income` = income_10k,
    `Hundres of people per km2` = density_100,
    `Average Household Size` = average_household_size,
    `Not suitable and major repairs needed` = unsuitable_major_repairs_pct,
    `2016 to 2021 population growth` = population_growth_pct
  ) %>% 
  summarise(
    across(
      c(
        `Percent Renting`,
        `Median Age`,
        `Median Total Income`,
        `Hundres of people per km2`,
        `Average Household Size`,
        `Not suitable and major repairs needed`,
        `2016 to 2021 population growth`
      ),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q3 = ~ quantile(.x, 0.75, na.rm = TRUE)
      )
    )
  )  %>% 
  pivot_longer(
    everything(),
    names_to = c("Variable", ".value"),
    names_pattern = "(.*)_(mean|median|q1|q3)$"
  )


knitr::kable(
  predictor_summary,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  caption = "Summary of candidate predictors"
) %>%
  kableExtra::kable_styling(
    latex_options = "scale_down",
    font_size = 7
  )
```

::: small-text
We run three types of regression to model the affordability metrics.

1.  OLS: modeling the mean predicted value of affordability
2.  Fractional Logit: treats affordability as an expected proportion, ensuring all fitted values are between 0 and 1
:::

## Best way to model

::: small-text
- We compared OLS and fractional logistic regression and found very little difference in predictive performance.
:::

```{r}
cv_rmse_log_density <- data.table::fread(
  "H:/github_repos/Assignment_MMAH/data/processed/rmse_log_results.csv"
)


knitr::kable(
  cv_rmse_log_density,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  caption = "10-fold cross-validation RMSE (percentage points)"
) %>%
  kableExtra::kable_styling(
    latex_options = "scale_down",
    font_size = 7
  )

```

::: small-text
In addition, we conducted sensitivity analyses:
:::

## Model Results

```{r ame_table}

frac_results <- data.table::fread(
  "H:/github_repos/Assignment_MMAH/data/processed/avg_marg_effects.csv"
)

frac_table <- frac_results %>%
  mutate(
    sig_30 = case_when(
      p_30 < 0.001 ~ "***",
      p_30 < 0.01  ~ "**",
      p_30 < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    
    sig_50 = case_when(
      p_50 < 0.001 ~ "***",
      p_50 < 0.01  ~ "**",
      p_50 < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    
    estimate_30 = sprintf(
      "%.3f%s",
      estimate_30,
      sig_30
    ),
    
    ame_30 = sprintf(
      "%.3f",
      ame_30
    ),
    
    p_30 = ifelse(
      p_30 < 0.001,
      "<0.001",
      sprintf("%.3f", p_30)
    ),
    
    estimate_50 = sprintf(
      "%.3f%s",
      estimate_50,
      sig_50
    ),
    
    ame_50 = sprintf(
      "%.3f",
      ame_50
    ),
    
    p_50 = ifelse(
      p_50 < 0.001,
      "<0.001",
      sprintf("%.3f", p_50)
    )
  ) %>%
  select(
    term,
    estimate_30,
    ame_30,
    estimate_50,
    ame_50,
  )

knitr::kable(
  frac_table,
  format = "latex",
  col.names = c(
    "Predictor",
    "30%+ Estimate",
    "30%+ AME",
    "50%+ Estimate",
    "50%+ AME"
  ),
  booktabs = TRUE,
  digits = 1,
  caption = "Fractional logistic regression results"
) %>%
  kableExtra::kable_styling(
    latex_options = "scale_down",
    font_size = 7
  )
```

::: small-text
- As we can see the estimates between the two outcomes are not substantially different.

  - A one-person increase in average household size is associated with an approximately **12 percentage-point increase** in the share of households experiencing affordability challenges.

  - A one-point increase in the percent of unsuitable/disrepair dwellings is associated with a **3 percentage-point increase** in the share of households experiencing affordability challenges.
:::

## Implications and other data sources:

::: small-text
- The CMHC and Ontario Government provide two additional data sources.
:::

```{r policy_table}
highest_affordability <- analytic_data |>
  filter(!is.na(affordability_30plus)) |>
  arrange(desc(affordability_30plus)) |>
  transmute(
    Municipality = GEO,
    `Household Size (2.6)` = average_household_size,
    `% Renter (22.8%)` = renter_pct,
    `30%+ (%)` = affordability_30plus,
    `Housing target status` =
      dplyr::coalesce(
        housing_target_status,
        "Not available"
      ),
    `2020 starts per 1,000` =
      starts_2020_per_1000_clean
  ) |>
  slice_head(n = 10)

knitr::kable(
  highest_affordability,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  caption = "Ten CSDs with the highest 30%+ affordability burden"
) %>%
  kableExtra::kable_styling(
    latex_options = "scale_down",
    font_size = 7
  )

```

::: small-text
- Among the ten CSDs with the highest affordability burden, only **Brampton and Oshawa** had more than 0.1 housing starts per 1,000 residents in the available 2020 CMHC data.

- Moving forward, only the City of Toronto has exceeded its housing target.
:::

## Limitations

::: small-text
There are many limitations to this study:

1.  There is no time element.
    1.  We're unable to capture the trend in affordability. Using cross sectional data prevents us from answering the very simple question of 'is affordability worse?'.
    2.  CMHC data only went as far back as 2020. Housing starts captures January 2020 starts and does not allow us to see a five or ten-year trend.
2.  The 2021 Census was affected by Covid-19

From the regression model, we could conclude that the improving these two policies would improve affordability:

1.  Repairing the existing housing stock
2.   Funding for more growth through Building Faster Fund
:::
