---
title: "MMAH Presentation"
author: "Kieran Shah"
editor: visual
format:
  revealjs:
    css: styles.css
    slide-number: true
    embed-resources: true
output:
  ioslides_presentation:
    widescreen: true
    self_contained: false
params:
  data_file: "data/processed/final_analytic_dataset.csv"
  output_dir: "eda/"
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150
)

knitr::opts_knit$set(
  root.dir = "H:/github_repos/Assignment_MMAH/"
)

library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(scales)
library(sf)
library(ggplot2)
library(scales)
library(knitr)
library(stringr)
library(leaflet)

analytic_file <- "H:/github_repos/Assignment_MMAH/data/processed/final_analytic_dataset.csv"

# Point this to the .shp file you downloaded
boundary_file <- "C:/Users/shahki/Downloads/lcsd000b21a_e/lcsd000b21a_e.shp"


# ------------------------------------------------------------
# 2. Read analytic data
# ------------------------------------------------------------

analytic_data <- fread(
  analytic_file,
  encoding = "Latin-1"
) %>%
  as_tibble() %>%
  dplyr::rename(
    population_2021 =
      `Population, 2021`,
    population_growth_pct =
      `Population percentage change, 2016 to 2021`,
    population_density =
      `Population density per square kilometre`,
    median_age =
      `Median age of the population`,
    average_household_size =
      `Average household size`,
    median_income_1p =
      `Median total income of one-person households in 2020 ($)`,
    renter_pct =
      `Renter`,
    unsuitable_major_repairs_pct =
      `'Not suitable' and 'major repairs needed'`,
    housing_target_status =
      `Housing Target Status`
  ) %>% 
  mutate(
    
    # Extract 7-digit Census Subdivision identifier
    CSDUID = str_sub(
      as.character(DGUID),
      -7
    ),
    
    # Calculate affordability measures
    affordability_30plus =
      100 *
      (`30% to less than 50%` + `50% or more`) /
      `Total - Shelter-cost-to-income ratio`,
    
    affordability_50plus =
      100 *
      `50% or more` /
      `Total - Shelter-cost-to-income ratio`,
    affordability_30plus_pct =
      100 * (`30% to less than 50%` + `50% or more`) / `Total - Shelter-cost-to-income ratio`,
    
    affordability_50plus_pct =
      100 * `50% or more` / `Total - Shelter-cost-to-income ratio`,
    
    log_population_density =
      log1p(population_density),
    
    # A municipality name in the CMHC join indicates actual CMHC coverage.
    # Rows without a CMHC match may currently contain zero after NA replacement;
    # restore these to NA so "not covered" is not interpreted as "zero starts".
    cmhc_covered =
      !is.na(municipality),
    
    starts_2020_cmhc =
      if_else(
        cmhc_covered,
        as.numeric(starts_2020),
        NA_real_
      ),
    
    starts_2021_cmhc =
      if_else(
        cmhc_covered,
        as.numeric(starts_2021),
        NA_real_
      ),
    
    starts_2020_per_1000_clean =
      if_else(
        cmhc_covered &
          !is.na(population_2021) &
          population_2021 > 0,
        1000 * starts_2020_cmhc / population_2021,
        NA_real_
      )
  )


```

## Introduction and Purpose

::: {.small-text}

Affordability is major concern across Ontario.

- A recent poll conducted by Angus Reid identified the cost of living and housing affordability as a the most important issue.

This presentation will seek to identify what issues are most associated with affordability.

A quick literature review suggests that the following could be causes.

- [Housing construction is not keeping up with population growth. Ontario's population grew much faster than anticipated between 2022 and 2023.](https://institute.smartprosperity.ca/sites/default/files/Ontarios-Need-for-1.7-Million-More-Homes-an-update.pdf?utm_source=chatgpt.com)

- [There are different regulatory barriers causing the increase in prices and decrease in affordability.](https://cdhowe.org/publication/benjamin-dachis-crunch-time-housing-ontario/?utm_source=chatgpt.com)

:::

## Data Sources {style="{.smaller}"}

::: {.small-text}

Statistics Canada provided the two main data sources for this presentation at the Census Subdivision level (CSD):

1.  `Table 98-10-0247-01` provided affordability variables.
2.  `Table 98-401-X2021021` is the summary of the predictor variables.

This presentation also explored the CMHC housing starts data from 2021.

And as a final bit of analysis, this presentation touched on the Ontario municipal housing target data.

Note, the two latter datasets covered many fewer municipalities.

:::

## EDA: Affordability Outcome Distribution {style="{.smaller}"}

```{r eda_1}


analytic_data %>%
  select(
    `Spending ≥30%` = affordability_30plus_pct,
    `Spending ≥50%` = affordability_50plus_pct
  ) %>%
  pivot_longer(
    everything(),
    names_to = "Measure",
    values_to = "Percent"
  ) %>%
  ggplot(aes(x = Percent)) +
  geom_histogram(
    bins = 25,
    boundary = 0
  ) +
  facet_wrap(
    ~ Measure,
    nrow = 1,
    scales = "free_y"
  ) +
  labs(
    x = "Households (%)",
    y = "Municipalities",
    title = "Distribution of housing affordability burden across Ontario"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

```

## Affordability Map

```{r mapping_slide}

csd <- st_read(
  boundary_file,
  quiet = TRUE
)


ontario_csd <- csd %>%
  filter(
    PRUID == "35"
  ) %>%
  mutate(
    CSDUID = as.character(CSDUID)
  )

map_data <- ontario_csd %>%
  left_join(
    analytic_data,
    by = "CSDUID"
  )

ap_data_leaflet <- map_data %>%
  st_make_valid() %>%
  
  # Use a projected coordinate system so tolerance is in metres
  st_transform(3347) %>%
  
  # Simplify boundaries: 500 m is usually plenty for an Ontario-wide map
  st_simplify(
    dTolerance = 500,
    preserveTopology = TRUE
  ) %>%
  
  # Leaflet requires longitude/latitude
  st_transform(4326)

pal_30 <- colorNumeric(
  palette = "magma",
  domain = map_data$affordability_30plus[
    !is.na(map_data$affordability_30plus)
  ],
  na.color = "#E5E5E5"
)

leaflet(ap_data_leaflet,
        height = "500px") %>%
  addPolygons(
    fillColor = ~pal_30(affordability_30plus),
    fillOpacity = 0.85,
    color = "white",
    weight = 0.4,
    label = ~paste0(
      CSDNAME, ": ",
      round(affordability_30plus, 1), "%"
    )
  ) %>%
  addLegend(
    position = "topright",
    pal = pal_30,
    values = ap_data_leaflet$affordability_30plus[
      !is.na(ap_data_leaflet$affordability_30plus)
    ],
    title = "Households<br>spending<br>≥30%"
  )
```

## EDA Predictors

```{r eda_pred}

univariate_plot_data <- analytic_data  %>% 
  transmute(
    affordability_30plus,
    `Renter share (%)` = renter_pct,
    `Median age (years)` = median_age,
    `Median income ($)` = median_income_1p,
    `Log population density` = log_population_density,
    `Average household size` = average_household_size,
    `Unsuitable + major repairs (%)` = unsuitable_major_repairs_pct,
    `Population growth 2016–2021 (%)` = population_growth_pct
  ) %>% 
  pivot_longer(
    cols = -affordability_30plus,
    names_to = "Predictor",
    values_to = "Value"
  ) %>% 
  filter(
    !is.na(affordability_30plus),
    !is.na(Value)
  )

univariate_correlations <- univariate_plot_data  %>% 
  group_by(Predictor)  %>% 
  summarise(
    r = cor(
      Value,
      affordability_30plus,
      use = "complete.obs"
    ),
    .groups = "drop"
  ) %>% 
  mutate(
    facet_label = paste0(
      Predictor,
      "\nr = ",
      sprintf("%.2f", r)
    )
  )

univariate_plot_data <- univariate_plot_data  %>% 
  left_join(
    univariate_correlations,
    by = "Predictor"
  )

ggplot(
  univariate_plot_data,
  aes(
    x = Value,
    y = affordability_30plus
  )
) +
  geom_point(
    alpha = 0.45,
    size = 1.2
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 0.7
  ) +
  facet_wrap(
    ~ facet_label,
    scales = "free_x",
    ncol = 3
  ) +
  labs(
    x = NULL,
    y = "Households spending ≥30% (%)",
    title = "Univariate relationships with housing affordability burden"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(
      face = "bold",
      size = 9
    ),
    plot.title = element_text(
      face = "bold"
    )
  )

```

## Modelling Approach {style="{.smaller}"}

::: {.small-text}

We consider the following variables as predictors when modelling our 30% and 50% affordability thresholds where we have 409 complete rows after missing cells are removed.

1.   The percent of the population renting
2.  The median Age
3.  The average household size
4.  The population growth between 2016 and 2021
5.  The median income
6.  The percent of the population with unsuitable repairs on their dwelling
7.  The density, and then the log of the density of the CSD

:::

::: {.small-text}

We run three types of regression to model the affordability metrics.

1.  OLS: modeling the mean predicted value of affordability
2.  Fractional Logit: treats affordability as an expected proportion, ensuring all fitted values are between 0 and 1

:::

## Best way to model

::: {.small-text}
- We compared OLS and fractional logistic regression and found very little difference in predictive performance.
:::

```{r}
cv_rmse_log_density <- data.table::fread(
  "H:/github_repos/Assignment_MMAH/data/processed/rmse_log_results.csv"
)

knitr::kable(
  cv_rmse_log_density,
  format = "html",
  digits = 2,
  table.attr = 'class="compact-kable"',
  caption = "10-fold cross-validation RMSE (percentage points)"
)

```

::: {.small-text}
In addition, we conducted sensitivity analyses:

1. Household-weighted models
2. Models including housing starts, which substantially reduced the available sample
:::

## Model Results

```{r ame_table}

frac_results <- data.table::fread(
  "H:/github_repos/Assignment_MMAH/data/processed/avg_marg_effects.csv"
)

frac_table <- frac_results %>%
  mutate(
    sig_30 = case_when(
      p_30 < 0.001 ~ "***",
      p_30 < 0.01  ~ "**",
      p_30 < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    
    sig_50 = case_when(
      p_50 < 0.001 ~ "***",
      p_50 < 0.01  ~ "**",
      p_50 < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    
    estimate_30 = sprintf(
      "%.3f%s",
      estimate_30,
      sig_30
    ),
    
    ame_30 = sprintf(
      "%.3f",
      ame_30
    ),
    
    p_30 = ifelse(
      p_30 < 0.001,
      "<0.001",
      sprintf("%.3f", p_30)
    ),
    
    estimate_50 = sprintf(
      "%.3f%s",
      estimate_50,
      sig_50
    ),
    
    ame_50 = sprintf(
      "%.3f",
      ame_50
    ),
    
    p_50 = ifelse(
      p_50 < 0.001,
      "<0.001",
      sprintf("%.3f", p_50)
    )
  ) %>%
  select(
    term,
    estimate_30,
    ame_30,
    estimate_50,
    ame_50,
  )

knitr::kable(
  frac_table,
  format = "html",
  col.names = c(
    "Predictor",
    "30%+ Estimate",
    "30%+ AME",
    "50%+ Estimate",
    "50%+ AME"
  ),
  align = c(
    "l",
    "r",
    "r",
    "r",
    "r"
  ),
  digits = 2,
  table.attr = 'class="compact-kable"',
  caption = "Fractional logistic regression results"
)
```

::: {.small-text}

- As we can see the estimate between the two outcomes are substantially different. The two most important variables are average household size and unsuitable repairs.

  - A one-person increase in average household size is associated with an approximately 12 percentage-point increase in the share of households experiencing affordability challenges.

  - A one-point increase in the percent of unsuitable/disrepair dwellings is associated with a 3 percentage-point increase in the share of households experiencing affordability challenges.

:::

## Implications and other data sources: 

::: {.small-text}

- The CMHC and Ontario Government provide two additional data sources.

:::

```{r policy_table}
highest_affordability <- analytic_data |>
  filter(!is.na(affordability_30plus)) |>
  arrange(desc(affordability_30plus)) |>
  transmute(
    Municipality = GEO,
    `Household Size (2.6)` = average_household_size,
    `% Renter (22.8%)` = renter_pct,
    `30%+ (%)` = affordability_30plus,
    `Housing target status` =
      dplyr::coalesce(
        housing_target_status,
        "Not available"
      ),
    `2020 starts per 1,000` =
      starts_2020_per_1000_clean
  ) |>
  slice_head(n = 10)

knitr::kable(
  highest_affordability,
  format = "html",
  table.attr = 'class="compact-kable"',
  digits = 1,
  caption = "Ten CSDs with the highest 30%+ affordability burden"
)
```

::: {.small-text}

- In 2020, only Brampton and Oshawa had more than .1 housing starts per 1000 households.

- Moving forward, only the City of Toronto has exceeded its housing target.

:::

## Limitations

::: {.small-text}

There are many limitations to this study:

1.  There is no time element.
    1.  We're unable to capture the trend in affordability. Using cross sectional data prevents us from answering the very simple question of 'is affordability worse?'.
    2.  CMHC data only went as far back as 2020. Housing starts captures January 2020 starts and does not allow us to see a five or ten-year trend.
2.  The 2021 Census
    1.  2020 and 2021 were affected Covid-19
        1.  Before tax income may be affected because there were temporary layoffs, coupled with government support.

:::
